### 2014-10-28 contract-language.tex
contract language

The syntax of our core contract language is shown in Fig.
~\ref{fig:contract-lang}. The language is based on first-order logic (FOL), and
admits prenex quantification over typed effect variables. We use a special
effect variable ($\cureff$) to denote the effect of \emph{current operation} -
the operation for which a contract is being written. The type of an effect is
simply the name of the operation (eg: \cf{withdraw}) that engendered the
effect. We admit disjuntion in types to let an effect variable range over
multiple operation names.

Quantifier-free propositions in our contract language are conjunctions,
disjunctions and implications of predicates expressing relations between pairs
of effect variables. Syntactic class of relations is seeded with primitive
$\visZ$, $\soZ$, and $\sameobjZ$ relations, and also admits derived relations
that are expressible as union, intersection, or transitive
closure\footnote{Strictly speaking, $R^{+}$ is not the transitive closure of
$R$, as transitive closure is not expressible in FOL. Instead, $R^{+}$ in our
language denotes \emph{a} superset of transitive closure of $R$. Formally,
$R^{+}$ is any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow R'(x,z)$} of
primitive relations. Commonly used derived relations are the \emph{same object
session order} ($\small \sooZ = \soZ ~\cap~ \sameobjZ$), the \emph{happens
before order} ($\small \hbZ = (\soZ ~\cup~ \visZ)^+$), and the per-object
happens-before order ($\small \hboZ = (\sooZ ~\cup~ \visZ)^+$).

@@

Similarly, a store that offers per-object causal consistency is captured by the
contract:

@@

This ability to represent store semantics and application-level consistency in
the same language is vital to contract classification. Observe that out
contract language does not incorporate real (i.e., wall-clock) time. Hence, the
contract language cannot describe store semantics based on real time such as
recency or bounded-staleness guarantees offered by certain
stores~\cite{Pileus}.

@@

Our goal is to classify the contracts, and map the corresponding operation to
the \emph{weakest} store level consistency. To this end, we need to be able to
compare the strengths of contracts. Let $\cv_{op}$ be a contract for a
particular operation $op$, and $\cv_{st}$ capture a particular store
consistency level. We would like to determine whether $op$ can be \emph{safely
discharged} at the store consistency level $\cv_{st}$ such that the resulting
execution does not violate $\cv_{op}$.

Since our contracts represent axiomatic definition of program executions, let
$\Mod{\cv}$ be the set of all executions under which $\cv$ is satisfied. If
every execution $\E \in \Mod{\cv_{st}}$ is also a member of $\Mod{\cv_{op}}$,
then $op$ can be safely discharged under the store consistency level
$\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq \Mod{\cv_{op}}$. This is the
model-theoretic consequence relation, written as $\cv_{st} \models_m \cv_{op}$.

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}. Due to the completeness theorem~\cite{completeness},
$\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st} \Rightarrow \cv_{op}$.
Due to the decidability of our contract language this implication check is
automatically discharged with the help of a theorem prover. In this case, we
say that $\cv_{op}$ is \emph{weaker than} $\cv_{st}$ (written $\cv_{op} \le
\cv_{st}$). Formally,

\begin{definition}
An contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-1em}

\noindent where $\Delta$ captures assumptions about the nature of primitive
relations, such as $\Rvis$ and $\Rso$ are irreflexive, and the happens-before
relation $\hbZ$ is acyclic, preventing thin-air reads. \KC{Ideally, we need to
list all the axioms somewhere.}

@@

While the store semantics offered by commercial and research data stores vary
widely, for our exposition, we identify three particular consistency levels --
eventual, causal and strong, commonly offered by many distributed store with
tunable consistency, with increasing overhead in terms of latency and
availability. Indeed, the techniques presented here can be extended to other
consistency stratifications in a straight forward manner. We assign each
application-level contract into one of these following classes:

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can be
satisfied as long as the client can reach at least one replica. For example,
\cf{deposit} is a highly-available operation. While eventually consistent data
store typically offer \emph{basic} eventual consistency with all possible
anomalies, we assume that our store provides stronger semantics that remains
highly-available~\cite{BailisHAT,COPS}; the store always exposes causal cut of
the updates. This semantics can be formally captured as:

\vspace{-1em}
\begin{smathpar}
\ecc = \forall a,b. \hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}

\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
contracts are required to see a causally consistent snapshot of the object
state, including the actions performed on the same session.  The latter
requirement entails that if two operations $o_1$ and $o_2$ from the same
session are applied to two different replicas $r_1$ and $r_2$, the second
operation cannot be discharged until the effect of $o_1$ is merged with $r_2$.
We assume that causality is only tracked through operations on the same object;
two operations in the same session but on different objects are considered
causally unrelated under this definition. Stores typically avoid tracking
causaility across objects to mitigate the overhead due to unintended causality.
The corresponding store semantics is captured by the contract $\ccc$ presented
in Section~\ref{sec:store_sem}.

\item \textbf{Strong Consistency}: Strongly consistent operations may be block
indefinitely under network partitions. An example is the total-order contract
on \cf{withdraw} operation. The corresponding store semantics is captured by
the contract $\scc$ presented in Section~\ref{sec:store_sem}.

@@

Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves totally
ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$. This
concurs with the intuition that any contract satisfiable under $\ecc$ or $\ccc$
is satisfiable under $\scc$ , and any contract that is satisfiable under $\ecc$
is satisfiable under $\ccc$. Nonetheless, we determine the class of a contract
based on the \emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$)
required to satisfy the contract. The classification scheme is presented
formally as rules in Figure~\ref{sem:classify}.

@@

The rules to classify contracts into availability classes are straightforward.
The remaining rule judges a contract as {\sf WellFormed} if and only if it can
be satisfied under strong consistency. The rest are reported as errors.

>>>>

The syntax of our core contract language is shown in Fig.
~\ref{fig:contract-lang}. The language is based on first-order logic (FOL), and
admits prenex universal quantification over typed effect variables. We use a special
effect variable ($\cureff$) to denote the effect of \emph{current operation} -
the operation for which a contract is being written. The type of an effect is
simply the name of the operation (eg: \cf{withdraw}) that induced the
effect. We admit disjuntion in types to let an effect variable range over
multiple operation names.

Quantifier-free propositions in our contract language are conjunctions,
disjunctions and implications of predicates expressing relations between pairs
of effect variables. The syntactic class of relations is seeded with primitive
$\visZ$, $\soZ$, and $\sameobjZ$ relations, and also admits derived relations
that are expressible as union, intersection, or transitive
closure\footnote{Strictly speaking, $R^{+}$ is not the transitive closure of
$R$, as transitive closure is not expressible in FOL. Instead, $R^{+}$ in our
language denotes \emph{a} superset of transitive closure of $R$. Formally,
$R^{+}$ is any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow R'(x,z)$} of
primitive relations. Commonly used derived relations are the \emph{same object
session order} ($\small \sooZ = \soZ ~\cap~ \sameobjZ$), the \emph{happens
before order} ($\small \hbZ = (\soZ ~\cup~ \visZ)^+$), and the per-object
happens-before order ($\small \hboZ = (\sooZ ~\cup~ \visZ)^+$).

@@

\noindent Similarly, a store that offers per-object causal consistency is captured by the
contract:

@@

\noindent This ability to represent store semantics and application-level consistency in
the same language is vital to contract classification. Observe that out
contract language does not incorporate real (i.e., wall-clock) time. Hence, the
contract language cannot describe store semantics based on real time such as
recency or bounded-staleness guarantees offered by certain
stores~\cite{Pileus}.

@@

Our goal is to classify contracts, and map the operation they describe to
the \emph{weakest} store-level consistency level that  nonetheless
satisfies the contract's constraints. To this end, we need a mechanism to 
compare the ``strength'' of a contract. Let $\cv_{op}$ be a contract for a
particular operation $op$, and $\cv_{st}$ capture a particular store
consistency level. We would like to determine whether $op$ can be
\emph{safely discharged} at the store consistency level $\cv_{st}$ such that
the resulting execution does not violate $\cv_{op}$.

Since our contracts represent axiomatic definition of program executions, let
$\Mod{\cv}$ be the set of all executions under which contract $\cv$ is satisfied. If
every execution $\E \in \Mod{\cv_{st}}$ is also a member of $\Mod{\cv_{op}}$,
then $op$ can be safely discharged under the store consistency level
$\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq \Mod{\cv_{op}}$. This is the
model-theoretic consequence relation, written as $\cv_{st} \models_m \cv_{op}$.

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}.  Completeness of the logic~\cite{completeness} allows
us to claim that $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st} \Rightarrow \cv_{op}$.
Due to the decidability of our contract language, this implication check is
automatically discharged with the help of a theorem prover. In this case, we
say that $\cv_{op}$ is \emph{weaker than} $\cv_{st}$ (written $\cv_{op} \le
\cv_{st}$).

@@

While the store semantics offered by commercial and research data stores
vary widely, for our purposes, we identify three particular consistency
levels -- eventual, causal, and strong, commonly offered by many distributed
stores with tunable consistency, with increasing overhead in terms of
latency and availability. Indeed, the techniques presented here can be
extended to other consistency stratifications. We assign each
application-level contract into one of these following classes:

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can be
satisfied as long as the client can reach at least one replica. For example,
\cf{deposit} is an eventually consistent operation. While eventually consistent data
store typically offer \emph{basic} eventual consistency with all possible
anomalies, we assume that our store provides stronger semantics that remain
highly-available~\cite{BailisHAT,COPS}; the store always exposes a causal cut of
the updates. This semantics can be formally captured in terms of the
following contract definition:
\vspace{-1em}
\begin{smathpar}
\ecc = \forall a,b. \hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}
\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
  contracts are required to see a causally consistent snapshot of the object
  state, including the actions performed on the same session.  The latter
  requirement entails that if two operations $o_1$ and $o_2$ from the same
  session are applied to two different replicas $r_1$ and $r_2$, the second
  operation cannot be discharged until the effect of $o_1$ is merged with
  $o_2$ in both $r_1$ and $r_2$.  We assume that causality is only tracked
  through operations on the same object; two operations in the same session
  but on different objects are considered causally unrelated under this
  definition. Stores typically avoid tracking causality across objects to
  mitigate overheads when causality tracking is unnecessary.  The corresponding
  store semantics is captured by the contract $\ccc$ presented in
  Section~\ref{sec:store_sem}.

\item \textbf{Strong Consistency}: Strongly consistent operations may block
  indefinitely under network partitions. An example is the total-order
  contract on \cf{withdraw} operation. The corresponding store semantics is
  captured by the contract $\scc$ presented in Section~\ref{sec:store_sem}.

@@

\noindent Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are
themselves totally ordered with respect to the $\le$ relation: $\ecc \le
\ccc \le \scc$. This concurs with the intuition that any contract
satisfiable under $\ecc$ or $\ccc$ is satisfiable under $\scc$ , and any
contract that is satisfiable under $\ecc$ is satisfiable under
$\ccc$. Nonetheless, we determine the class of a contract based on the
\emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$) required to
satisfy the contract. The classification scheme is presented formally 
in Figure~\ref{sem:classify}; here, $\Delta$ captures assumptions about the nature of primitive
relations, such as $\Rvis$ and $\Rso$ are irreflexive, and the happens-before
relation $\hbZ$ is acyclic, preventing thin-air reads. \KC{Ideally, we need to
list all the axioms somewhere.}  The following definition thus precisely
captures our intuition about contract strength in terms of these rules:

\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-1em}

@@

%% The rules to classify contracts into availability classes are straightforward.
%% The remaining rule judges a contract as {\sf WellFormed} if and only if it can
%% be satisfied under strong consistency. The rest are reported as errors.

### 2014-10-28 intro.tex
contract language

The rest of the paper is organized as follows. Section~\ref{sec:motivation}
motivates \name and introduces the language through a series of examples.
Section~\ref{sec:lang} describes the contract language, and
Section~\ref{sec:opsem} presents an operational semantics.
Section~\ref{sec:txns} introduces transaction contracts and classification.
Section~\ref{sec:impl} describes the implementation and goes into detail about
the optimizations needed to make the system practical.
Section~\ref{sec:results} discusses experimental evaluation.
Section~\ref{sec:related} and~\ref{sec:concl} present related work and
conclusions.

>>>>

The rest of the paper is organized as follows. The next section describes
the system model. Section~\ref{sec:motivation} describes the challenges in
programming with eventually consistent data stores, and introduces
\name\ contracts as a proposed solution to overcome these issues.
Section~\ref{sec:lang} provides more details the contract language.
Section~\ref{sec:opsem} presents an operational semantics.
Section~\ref{sec:txns} introduces transaction contracts and classification.
Section~\ref{sec:impl} describes the implementation and provides details 
about the optimizations needed to make the system practical.
Section~\ref{sec:results} discusses experimental evaluation.
Section~\ref{sec:related} and~\ref{sec:concl} present related work and
conclusions.

### 2014-10-28 motivation.tex
edits
motivation.tex
contract language

In this section, we motivate the need and benefit of declarative reasoning for
eventually consistent systems.  We begin with a description of our system and
programming model, and introduce the primitive relations that constitute the
essential part of our contract language.

\subsection{System Model}
\label{sec:sysmod}

@@

\subsection{Example}

@@

Although it is easy to understand the reasons behind the occurrence of the
aforementioned negative balance and missing update anomalies, finding the
appropriate fixes is not readily apparent. Making \cf{getBalance} a strongly
consistent operation is definitely sufficient to avert anomalies, but is it
really necessary? Given the cost of enforcing strong
consistency~\cite{DynamoDB, Pileus}, it is preferable to avoid unless there are
no viable alternatives.  Exploring the space of these alternatives requires
understanding the subtle differences in semantics of various kinds of weak
consistency alternatives.

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract or specification over \cf{withdraw}:

\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}

Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect which was emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

For \cf{getBalance}, we construct the following contract:

\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \wedge~ (\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff}
\end{array}
\end{smathpar}

The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an
effect that was emitted either by a \cf{deposit} or a \cf{withdraw} operation.
If a \cf{withdraw} $b$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $a$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

Let us consider the last line in the above contract. Our specification language
provides operators to for composing relations. Thus, the syntax $(R_1 \cap
R_2)(a,b)$ is equivalent to $R_1(a,b) \wedge R_2(a,b)$. The last line of the
above contract says that if a \cf{deposit} or a \cf{withdraw} precedes a
\cf{getBalance} operation in session order, and is applied on the same object
as the \cf{getBalance} operation, then it must be the case that the
\cf{getBalance} operation witnesses the effects of the preceding operations.

>>>>

\section{System Model}
\label{sec:sysmod}

@@

@@

Although it is easy to understand the reasons behind the occurrence of the
aforementioned negative balance and missing update anomalies, finding the
appropriate fixes is not readily apparent. Making \cf{getBalance} a strongly
consistent operation is definitely sufficient to avert anomalies, but is it
really necessary? Given the cost of enforcing strong
consistency~\cite{DynamoDB, Pileus}, it is preferable to avoid it unless
there are no viable alternatives.  Exploring the space of these alternatives
requires understanding the subtle differences in semantics of various kinds
of weak consistency alternatives.

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract or specification over \cf{withdraw}:
\vspace{-.5em}
\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}
\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect which was emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

\noindent For \cf{getBalance}, we construct the following contract:
\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \wedge~ (\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff}
\end{array}
\end{smathpar}
\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an
effect that was emitted either by a \cf{deposit} or a \cf{withdraw} operation.
If a \cf{withdraw} $b$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $a$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

Our specification language provides operators to compose
relations. Thus, the syntax $(R_1 \cap R_2)(a,b)$ is equivalent to $R_1(a,b)
\wedge R_2(a,b)$. The last line of the above contract says that if a
\cf{deposit} or a \cf{withdraw} operation precedes a \cf{getBalance} operation in
session order, and is applied on the same object as the \cf{getBalance}
operation, then it must be the case that the \cf{getBalance} operation
witnesses the effects of the preceding operations.
